Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-19 paper.qmd
finished types of papers section

@@

## Types of Uncertainty Visualisation Papers
Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

There are some papers that suggest a novel plot that does not serve to fix a specific visualisation issue. These papers suggest ways to include even *more* information in a visualisation, and shift the information overload from the visualisation author to the viewer. Clever methods of combining several types of information, such as VSUPs or tours, can better help us understand how our brains process information in a graphic, but visualisations that simply seek to display as much information as possible fail to understand that visualisations are primarily as a tool to summarise information. Poor understanding of our goals in generating a visualisation will not be remedied by overloading the viewer.

The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments. There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions. These questions can be about specific statistics (e.g. What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g. How much do you trust this estimate?).
- Perceptual Tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic. 
- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different

Heuristic checks: do people have this bias when looking at this plot?
Meta papers: Uncertainty visualisation experiments should actually be done like this.
Qualatative analysis: Look at this visualisaiton for this purpose, tell us what you are looking at and why.
Can people get X: Draw/look at this uncertainty viusalisation, what is the mean of this distribution?

Each of these types of papers may have distinct goals in their visualisation, but their problems and conflicting results share a common source. Some of these issues come from issues deeper than the visualisation field (e.g. the lacking definitions of uncertainty) while some are specific to visualisation (e.g. the difficulty in identifying the signal of a graphic). Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.

@@

- perceptual tasks need to know what you are uncertain about
- perceptual tasks don't consider

@@

- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
- Why are spatial and temporal data considered special? What is the statistical element that remains after they have been away? Inexchangeability. Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
  - 
- Yes the data type is relevant but its not the most important thing
  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
  - screenshot from these three papers, the connecting factor is the question, not the data type
  - The data type: spatial, temporal, etc, usually is mentioned as the 
  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?". This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017

>>>>

There are several frameworks that the uncertainty visualsiation problem is approached from. One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved. Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.

@@

## Types of Uncertainty Visualisation Papers
There are two broad categories of uncertainty visualisation experiments, depending on what they believe the "problem" in uncertainty visualisation to be.

The first type of paper believes uncertainty to be fundamentally differently to other variables due to psychological elements of uncertainty, such as risk aversion. These papers seek to investigate uncertainty visualisaiton by assessing which visualisation methods facilitate these "risk" related biases and check that perceptual tasks are not impacted by the fact that "uncertainty" is what is being visualised. At least this is often how the motivation is described. Due to the fact that uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, these papers are all over the place in terms of methodology. Since it is unclear exactly *what* they are testing for, since it is unclear *how* uncertainty information is suppoesdly different from any other information, these papers are poorly justified and attempt to test for the secondary effects of a psychological outcome that has actually never been justified or shown to exist.

The second type of paper perceives uncertainty to be the same as any other variable, and therefore the issue facing uncertainty visualisation is actually a dimensionality problem. These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation. Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance. This is not to say *every* visualisation method that sees it as a dimensionality problem is bad. Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic. However this is because visualisations that simply seek to display as much information as possible fail to understand visualisation are a tool to summarise information using visual features, and once it fails to summarise information it is no longer a good visualisation. Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.

In addition to the way the papers *see* the uncertainty visualisation problem, these papers can also be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments. There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:  
- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.  
- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.  
-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions. These questions can be about specific statistics (e.g. What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g. How much do you trust this estimate?).  
- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation. 
- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.

These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories. These distinctions are important because each have their own issues in how they relate to uncertainty more broadly. Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.

@@

@@

- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
- Why are spatial and temporal data considered special? What is the statistical element that remains after they have been away? Inexchangeability. Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
- perceptual tasks need to know what you are uncertain about.
- Yes the data type is relevant but its not the most important thing
  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
  - screenshot from these three papers, the connecting factor is the question, not the data type
  - The data type: spatial, temporal, etc, usually is mentioned as the 
  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?". This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
